\subsection{Polyregular functions}

\paragraph{Words and languages}  
Let $\Sigma$ be an alphabet, we denote by $\Sigma^*$
 the set of finite words over $\Sigma$. The empty word is written $\epsilon$. If $w\in \Sigma^*$, $w[i]$ is the $i^{th}$ letter of the word $w$.

The powerset of $E$ is denoted by $\mathbf 2^E$.

If $i, j \in \mathbb{N}$, $[i,j]$ denotes the set $\set{m \in \mathbb{N} |\ i\leq m\leq j}$. In particular, if $ i>j$ then $[i,j]=\emptyset$. 

\subsubsection{$1$-pebble transducers}
A $1$-pebble transducer, (usually known as a two-way transducer) of input alphabet $\Sigma$ and output alphabet $\Gamma$ is a two way automaton (meaning that it has a reading head, called here a pebble, which can scan the word on both directions) which reads words over $\Sigma^*$, and has the ability to output words over $\Gamma^*$ on every transition. 
The output of a one pebble transducer is the concatenation of the outputs of the transitions it took along a run.

\begin{definition}\label{def:1pebble}
A $1$-pebble transducer is a tuple $(\Sigma,\Gamma, Q, q_I, q_F, \delta)$, which consists of:
\begin{itemize}
\item a finite input alphabet $\Sigma$ and a finite output alphabet $\Gamma$; 
\item a finite set of states $Q$;
\item two designated states $q_I$ and $q_F$: the initial and final one;
\item  a transition function of type 
$$\delta : Q\times \Sigma\cup\{\vdash,\dashv\} \to Q\times\set{\rightarrow,\leftarrow}\times\Gamma^*$$
The symbols  $\vdash$ and $\dashv$ are the endmarkers of the word. 
\end{itemize}
We assume that the transducer can only move to the right when it is on the left endmarker $\vdash$, and only to the left when  it is on the right endmarker $\dashv$. We also assume that the endmarkers don't output anything, meaning  $\delta(Q\times\set{\vdash,\dashv})\subseteq Q\times\set{\rightarrow,\leftarrow}\times \set{\epsilon}$.
 \end{definition}


Let us define the behavior of the transducer over an input word $w\in\Sigma^*$. The transducer reads the word ${\vdash} w{\dashv}$; and we denote by $\Sigma_{\vdash\dashv}$ the set  $\Sigma\cup\set{\vdash,\dashv}$. \emph{A configuration} is a pair $(q,i)$ where $q$ is the control state and $i$ the position of the pebble on ${\vdash} w{\dashv}$. 

Let $(p,i)$ be a configuration and suppose that $\delta(p, {\vdash} w{\dashv}[i])=(q,d,\gamma)$. \emph{The successor of $(p,i)$} is the configuration $(q,j)$
where:
\begin{align*}
 j=i+1 &\qquad \text{ if }\qquad d=\ \rightarrow\\
 j=i-1 &\qquad \text{ if }\qquad d=\ \leftarrow 
\end{align*}
\emph{The output of $(p,i)$} is the word $\gamma$. \emph{A run} on $w$ is a sequence of configurations over $w$ related by a the successor relation defined above. The output of a run is the word obtained by concatenating the outputs of its configurations.

% such that $c_{i+1}$ is the successor of $c_i$. \emph{The output} of a run is the word $(\mathsf{out}(c_i))_{0\leq i\leq n}$. 

The initial configuration is $(q_I, 0)$ and the final configuration is $(q_F,|w|+1)$. The automaton accepts $w$ if there is an accepting run,
i.e. a run where the first configuration is initial, the last one is final, and no other configuration is final. The accepting run, if it exists, is unique, by determinism of the transition function.


\subsubsection{$k$-pebble transducers}

In the literature (see \textit{e.g.}~\cite{Bojanczyk18}), a $k$-pebble transducer is a transducer with $k$ reading heads. The movement of these heads is subject to a stack discipline: only the pebble on top of the stack can move.
In this paper, we will work with a different yet equivalent definition of $k$-pebble transducer. Here a $k$-pebble transducer is a collection of $k$ $1$-pebble transducers. The idea is that the transducer number $k$ can, along its run, call transducer $k-1$ to run over \emph{its} current configuration. Then transducer $k-1$ can itself call transducer $k-2$ to run over its current configuration, and so on.
\begin{center}
Picture
\end{center}
Let $\Sigma$ be a finite alphabet and let $S$ be a finite set of \emph{labels}, we define the alphabet of labelled letters $\Sigma(S)=\Sigma\times \mathbf 2^S$. For $S\cap T=\varnothing$, we identify the sets $(\Sigma(S))(T)$ and $\Sigma(S\cup T)$, as well as the sets $\Sigma(\varnothing)$ and $\Sigma$.
Let $a=(b,T)\in \Sigma(S)$, and let $s\in S$, then we denote by  $a(s)$ the letter $(b,T\cup\set{s})$.
%Let $k, i\in \mathbb{N}$.
%We set $\Sigma_{[i,k]}=\Sigma\cup\Sigma\times \mathbf 2^{[i,k]}$.  
%In the alphabet $\Sigma_{[k, i+1]}$, we have the possibility to mark some letters of $\Sigma$ by the information of wich pebbles are under head. 

\begin{comment}
\begin{definition}%[$k$-pebble transducer]
A $k$-pebble transducer of input alphabet $\Sigma$ and output alphabet $\Gamma$ is a tuple $\Tt=(T_k,\dots,T_{1})$ such that for every $i\leq n$:
\begin{itemize}
\item  $T_i$ is 1-pebble transducer, whose set of states is $Q_i$;
\item  The input alphabet of $T_i$ is $\Sigma_{[i+1,k]}$;
\item  The output alphabet of $T_i$ is $\Gamma\cup\bigcup_{1\leq j <i } Q_j$.
\end{itemize} 
In particular, the input alphabet of $T_k$ is $\Sigma$ and the output alphabet of $T_1$ is $\Gamma$. We denote by $\Tt_i$ the transducer $T_i$. 
\end{definition}
\end{comment}

\begin{definition}%[$k$-pebble transducer]
  A $k$-pebble transducer of input alphabet $\Sigma$ and output alphabet $\Gamma$ is a tuple $\Tt=\tuple{T_1,\dots,T_{k}}$ such that for every $i\leq n$:
  \begin{itemize}
  \item  $T_i$ is a 1-pebble transducer, whose set of states is denoted $Q_i$;
  \item  The input alphabet of $T_i$ is $\Sigma(Q_{>i})\quad$ (with $Q_{>i}=\bigcup_{j\in [i+1,k]}Q_j$);
  \item  The output alphabet of $T_i$ is $\Gamma \cup [1,i-1]$.
  \end{itemize} 
  In particular, the input alphabet of $T_k$ is $\Sigma$ and the output alphabet of $T_1$ is $\Gamma$.%We denote by $\Tt_i$ the transducer $T_i$. 
  \end{definition}

%The alphabet $(\Gamma_{[l,m]})_{[m+1,n]}$ can be seen as the alphabet $\Gamma_{[l,n]}$. Indeed, every letter of $(\Gamma_{[l,m]})_{[m+1,n]}$ of the form $((a,e),f)$ where $e\subseteq [l,m]$ and  $f\subseteq [m+1,n]$ can be seen a a letter $(a,e\cup f)$ of $\Gamma_{[l,n]}$ and conversely.

For every $k$-pebble transducer $\Tt=\tuple{T_j}_{j\in[1,k]}$ and $i\in[1,k]$,  the sequence $\tuple{T_j}_{j\in[1,i]}$ can be seen as an $i$-pebble transducer, of input alphabet $\Sigma(Q_{>i})$ and output alphabet $\Gamma$. We denote this transducer by $\Tt_i$.
%Similarly, the sequence   $(T_j)_{j\in[i+1,k]}$ can be seen as a $(k-i)$-pebble transducer, of input alphabet $\Sigma$ and output alphabet $\Gamma \cup [1,i]$. We denote them respectively by $\Tt[1,i]$ and $\Tt[i+1,k]$.
 
 \medskip
\begin{terminology} 
 A $k$-pebble transducer $\Tt$ can be seen as the one-pebble transducer $T_k$, which outsources a part of the computation to the other one-pebble transducers $T_i$, $i<k$.
 Transducer $T_i$ can then itself call as subroutines the transducers of lower indexes, and so on.
 For this reason, we call \emph{the states of $\Tt$} the states of $T_k$, and \emph{the initial state of $\Tt$} the initial state of $T_k$. 
  \end{terminology}
  
  \smallskip
 Let us define the function realized by a $k$-pebble transducer. 
  
  \begin{definition}
We define, by induction on $k$, the function realized by a $k$-pebble transducer. The case $k=1$ has been treated in Definition~\ref{def:1pebble}. 

Consider a $k+1$ pebble transducer $\Tt=\tuple{T_1,\ldots,T_{k+1}}$, and let $Q_i$ be the set of states of $T_i$, for $i\in [1,k+1]$. By induction let $f_i:\Sigma(Q_{>i})^*\rightarrow \Gamma^*$ denote the transduction realized by $\Tt_i$, for $i\in [1,k]$.
Let us define the image of a word $w$ of $\Sigma^*$ by the transduction realized by $\Tt$:
\begin{itemize}
  \item Let $r=(q_j,p_j)_{j\in[1, n+1]}$ be the accepting run of $T_{k+1}$ over $w$ and $(\gamma_j)_{j\in[1, n]}$ be the outputs of the corresponding configurations.
  
  \item  For every $j\in[1,n]$, let $w_j$ be the word obtained from $w$ by replacing the letter $a$ at position $p_j$ by $a(q_j)$.
      For every $j\in[1,n]$, let $u_j$ be the word obtained from $\gamma_j$ by replacing each occurrence of a letter $i\in [1,k]$ by $f_i(w_j)$.
  
  \end{itemize}
  The image of $w$ by $\Tt$ is the word $u_1\cdots u_n$.
\end{definition}


\begin{example}
  \begin{center}
  Add an example.
  \end{center}
  \end{example}
  

\begin{definition}
  Let $f:\Sigma^*\rightarrow \Gamma^*$ be a word to word function. Its \emph{size function}, $|f|:\NN\rightarrow \NN$ is defined by $|f|(n)=\max_{|w|\leq n}|f(w)|$. For $k\in \NN$, we say that the function $f$ has \emph{degree $k$ size output} if $|f|(n)$ is equal to $\Oo(n^k)$ . Similarly, we say that $f$ has an output size of degree \emph{strictly} $k$ if $|f|(n)=\Theta(n^k)$. For $k=0,1,2,\ldots$, we use the usual terms bounded, linear, quadratic, \textit{etc}.
  If there exists $k$ such that $f$ has a degree $k$ size output, then we says that $f$ has polynomial size output.
  
\end{definition}

\begin{comment}

\begin{definition}
  We define, by induction on $k$ the function, realized by a $k$-pebble automaton. The case $k=1$ has been treated in Definition~\ref{def:1pebble}. 
  
  Consider a $k+1$ pebble transducer $\Tt$. We set $Q_i$ to be the set of states of $\Tt_i$.
  Let us define the image of a word $w$ of $\Sigma^*$ by the transduction realized by $\Tt$.
  
  \noindent Let $r=(q_j,\mathsf{pos}_j)_{j\in[1, n]}$ be the accepting run of $\Tt_{k+1}$ over $w$ and $(o_j)_{j\in[1, n]}$ be the outputs of the corresponding configurations.
  
  \noindent For every $q\in Q_i$, let $f^i_q:\Sigma_{[k,i+1]}^*\rightarrow \Gamma^*$ be the transduction realized by $\Tt[1,i]$, considering $q$ as its initial state.
   
   For every $j\in[1,n]$, let $w_j$ be the word obtained from $w$ by replacing the letter $a$ at position $\mathsf{pos}_j$ by $(a,k+1)$.
      For every $j\in[1,n]$, let $u_j$ be the word obtained from $o_j$ by replacing every state $q\in Q_i$ by $f^i_q(w_j)$.
      
  The image of $w$ by $\Tt$ is the word $(u_j)_{j\in[1,n]}$.
  \end{definition}

    
\end{comment}




\begin{proposition}[\cite{}]
  \label{prop:degree}
The function realized by a $k$-pebble transducer has degree $k$ size output.
\end{proposition}

\subsection{Transition monoids}
We present in this section a tool used to summarize the behavior of a one-pebble automaton, and called its transition monoid (resp. morphism). We map each word  $w$ to an element of the monoid which gives the following kind of information \textit{e.g}~: if the automaton enters the word from the left in state $q$, then it exits to the left in state $q'$, \textit{etc}.
Moreover, we will sometimes need to record information about the output produced in such a pass of the transducer; this information can be for instance the whole output (in which case the monoid is infinite) or information of the kind ``a letter $a$ has been produced at least once'' (here we recover finiteness).

   
\begin{definition}%[Transition monoid of a 1-pebble transducer]
Let $T$ be a 1-pebble automaton with set of states $Q$. Let $N$ be a monoid, and let $\star$ be its multiplication.

We define the transition monoid $M_N$ of $T$ as follows:
\begin{itemize}
\item its elements are functions of the form $f:Q\times\set{\rightarrow,\leftarrow}\to Q\times\set{\rightarrow,\leftarrow}\times N$;
\item the composition $\cdot$ is defined as follows. Let $f, g$ be two elements of $M_N$, $q\in Q$ and $d\in\set{\rightarrow, \leftarrow}$. We define the \emph{transition sequence} between 
$f$ and $g$ starting from $(q,d)$ and its \emph{output sequence} to be respectively the sequences $(q_i,d_i)_{i\in[0,n]}$ and  $(w_i)_{i\in[1,n]}$ satisfying the following conditions: 
\begin{itemize}
\item $(q_0,d_0)=(q,d)$;
\item $d_{n-1}=d_{n}$ and $d_i\neq d_{i+1}$ for every $i\in[1,n-2]$;
\item if $d_0=\rightarrow$ then for every even $i$, $f(q_i,d_i)=(q_{i+1},d_{i+1}, w_{i+1})$ and for every odd $i$, $g(q_i,d_i)=(q_{i+1},d_{i+1}, w_{i+1})$;
\item if $d_0=\leftarrow$ then for every even $i$, $g(q_i,d_i)=(q_{i+1},d_{i+1}, w_{i+1})$ and for every odd $i$, $f(q_i,d_i)=(q_{i+1},d_{i+1}, w_{i+1})$. 
\end{itemize}
We set $(f\cdot g) (q,d)$ to be $(q_n, d_n, w_1\star\dots\star w_n)$.
\end{itemize}
\end{definition}
We will mainly instantiate $M_N$ in the following three cases: 
\begin{enumerate}
\item $N$ is the monoid $\Gamma^*$ of words over $\Gamma$.
\item $N$ is the two element monoid $\mathbf 2$ with operation $\max$.
\item $N$ is the singleton monoid $\mathbf 1$.
\end{enumerate}
In the last case, the third component of the codomain of the elements of $M_{\mathbf 1}$ is useless, one can be see them as functions of type $Q\times\set{\rightarrow,\leftarrow}\to Q\times\set{\rightarrow,\leftarrow}$.

\begin{example}
\begin{center}
Show two boxes and their composition.

Give an example of transition sequence.
\end{center}
\end{example}

Let us show define the transition morphisms associated with a 1-pebble transducer.
\begin{definition}
Let $T=(\Sigma, \Gamma, Q, q_I, q_F, \delta)$ be a 1-pebble transducer. 
We define the morphism $\mu:(\Sigma_{\vdash\dashv})^*\to M_{\Gamma^*}$ as follows:
$$\text{For every } d\in \set{\rightarrow,\leftarrow}\qquad\mu(a)(q,d)=\delta(a,q)$$
Let $\Delta\subseteq \Gamma$. We define the morphism $\mu_\Delta:(\Sigma_{\vdash\dashv})^*\to M_{\mathbf 2}$ as follows. Let $1_\Delta:\Gamma^*\to \mathbf{2}$ be the morphism defined on letters as follows $1_\Delta(\gamma)=1$ if $\gamma\in \Delta$ and $1_\Delta(\gamma)=0$ otherwise. We set then for every $d\in\set{\rightarrow,\leftarrow}$: 

$$\text{If } \delta(a,q)=(q',d',w) \text{ then } \mu_\Delta(a)(q,d)=(q',d',1_\Delta(w))$$
We define the morphism $\mu_{\mathbf{0}}:(\Sigma_{\vdash\dashv})^*\to M_{\mathbf 0}$ as follows. For every $d\in\set{\rightarrow,\leftarrow}$,
$$\text{if } \delta(a,q)=(q',d',w) \text{ then }  \mu(a)(q,d)=(q',d') $$
\end{definition} 

